---
title: "Example 3: Necklace Sampler"
author: "Yara Skaf"
date: "1/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(ggrepel)
library(landmark)
library(tdaunif)
```

This vignette provides a clear example of a situation where the differences between the maxmin and lastfirst algorithms result in significant differences between the two landmark sets, and thus between the resulting covers.

Let $X$ be the data set shown below and equip $X$ with the standard Euclidean metric.

```{r dataSetup}
set.seed(0)
sample_circle_radius <- function(n, bins = 1L, sd = 0, r=1L) {
  theta <- sample_strat_segment(n, bins) * 2*pi
  res <- cbind(x = r*cos(theta), y = r*sin(theta))
  colnames(res) <- c("x","y")
  data.frame(add_noise(res, sd = sd))
}

# sample a few points around a large circle for the "necklace" outline
bigCircle <- sample_circle_radius(20, sd=0.05, r=10)

# add points sampled from smaller circles to use as the "beads" on the necklace
dat <- rbind(bigCircle,
             sample_circle_radius(100, sd=0.1) %>% mutate(y=y+10),
             sample_circle_radius(100, sd=0.1) %>% mutate(x=x+10),
             sample_circle_radius(100, sd=0.1) %>% mutate(x=x-10),
             sample_circle_radius(100, sd=0.1) %>% mutate(y=y-10)
             )

# plot the resulting data set
g <- ggplot(dat, aes(x, y)) +
  geom_point() +
  theme_minimal()
g
```



Observe that calculating the first few landmark points of $X$ gives different results with each procedure:


```{r}
# num <- 5L works fine, but any number 6L or higher seems to cause the error
# using seed_index=1L also gives the error for landmarks_lastfirst
num <- 4L
input <- as.matrix(dat)
lmk_mm <- landmarks_maxmin(input, num = num, seed_index = "minmax")
lmk_lf <- landmarks_lastfirst(input, num = num, seed_index = "firstlast")
```

```{r}
highlight_mm <- data.frame(dat[lmk_mm,] %>% cbind(name=c(1:length(lmk_mm)-1))) %>%
  mutate(name=paste0("l",name))
g_mm <- ggplot(dat, aes(x, y)) +
  geom_point() +
  geom_point(data=highlight_mm,
             aes(x=x, y=y),
             color='red',
             size=3) +
  geom_text_repel(data=highlight_mm, aes(label=name)) +
  labs(title = "MaxMin Landmarks") +
  theme_minimal()
g_mm

highlight_lf <- data.frame(dat[lmk_lf,] %>% cbind(name=c(1:length(lmk_lf)-1))) %>%
  mutate(name=paste0("l",name))
g_lf <- ggplot(dat, aes(x, y)) +
  geom_point() +
  geom_point(data=highlight_lf,
             aes(x=x, y=y),
             color='blue',
             size=3) +
  geom_text_repel(data=highlight_lf, aes(label=name)) +
  labs(title = "LastFirst Landmarks") +
  theme_minimal()
g_lf
```


